\documentclass[sigconf]{acmart}
\setcopyright{none}
\settopmatter{printacmref=false}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{xcolor}

\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

\copyrightyear{}
\acmYear{}
\acmConference[]{}{}{}
\acmISBN{}
\acmDOI{}

\begin{document}

\title{Assignment 2: Self-Attentive Sequential Recommendation (SASRec)}

\author{Mayuresh Khalane (s4714067)}
\affiliation{%
  \institution{Leiden University}
  \city{Leiden}
  \country{Netherlands}}
\email{m.j.khalane@umail.leidenuniv.nl}

\author{Xing He (s4707443)}
\affiliation{%
  \institution{Leiden University}
  \city{Leiden}
  \country{Netherlands}}
\email{x.he.8@umail.leidenuniv.nl}

\begin{abstract}
This report presents an implementation of SASRec trained and evaluated on
MovieLens~1M under full-ranking evaluation. A controlled ablation varies each
of four architectural dimensions --- hidden size, number of attention blocks,
number of heads, and sequence length --- one at a time against a shared
baseline, plus a combined configuration that sets all dimensions to their larger
values. Hidden size is the most impactful single dimension (+15.4\% NDCG@10
over baseline), while adding heads alone hurts performance when the per-head
dimension becomes too small (16 dims/head at H=64). The combined configuration
achieves a test Recall@10 of 0.2007 and NDCG@10 of 0.1032. Neither model
triggered early stopping at 50 epochs, suggesting the training budget was the
binding constraint rather than overfitting.
\end{abstract}

\maketitle


\section{Introduction}

Matrix factorization compresses a user's entire interaction history into a fixed
vector. Sequential recommendation models take a different approach: they treat
the history as an ordered sequence and learn which past items matter most for
predicting the next one, making temporal order a first-class feature rather than
something to be averaged away.

SASRec~\cite{kang2018self} addresses next-item prediction with a
Transformer-style architecture~\cite{vaswani2017attention}. Stacked multi-head
self-attention blocks process the interaction sequence with a causal mask, so
each position attends to all preceding items but not future ones. The model is
implemented in PyTorch and evaluated on MovieLens~1M~\cite{harper2015movielens}
using full-ranking NDCG@10, NDCG@20, Recall@10, and Recall@20.


\section{Data Preprocessing}

The MovieLens~1M dataset contains 1,000,209 ratings from 6,040 users on 3,706
movies. Each line has the format \texttt{userId::movieId::rating::timestamp}.

Ratings below 4.0 are discarded; only higher-rated films are kept as positive
interactions. The retained interactions are sorted by timestamp for each user,
giving a chronological item sequence. Users with fewer than five interactions
are removed, since the split described below requires at least three items and
very short sequences provide little sequential signal.

User and item IDs are remapped to contiguous integers starting from 1; index 0
is reserved for padding. For a user sequence of length $n$, the first $n-2$
items form the training sequence, item $n-1$ is used for validation, and item
$n$ is the test target. The complete interaction history per user is stored and
used at evaluation time to exclude already-seen items from the candidate set.

Training samples are constructed token-wise. For each non-padding position, the
input is the item at that position and the target is the next item. Sequences
are left-padded with zeros to a fixed length $L$; padding positions do not
contribute to the training loss.


\section{Model Architecture}

\subsection{Embeddings}

Each item is mapped to an $H$-dimensional vector via a learnable embedding table
of size $(N+1) \times H$, where index 0 is fixed to zero (padding). Position
embeddings of the same size are added elementwise:

\begin{align}
  \mathbf{x}_t = \mathbf{e}_{i_t} + \mathbf{p}_t, \quad t = 1, \dots, L.
\end{align}

Padding positions are zeroed out before the first block and after each
subsequent block. Residual connections can restore non-zero values at padding
slots even when the attention mask prevents information flow, so the explicit
zeroing is necessary.

\subsection{Self-Attention Blocks}

The model stacks $B$ identical blocks. Following Xiong et al.~\cite{xiong2020layer},
layer normalisation is applied before each sub-layer (pre-LN), which stabilises
gradient flow in deeper networks.

The self-attention sub-layer uses a causal mask: entry $(i, j)$ is blocked when
$j > i$, so position $i$ cannot attend to any future position. A key-padding
mask additionally prevents all positions from attending to padding tokens.

The point-wise feedforward sub-layer is:

\begin{align}
  \text{FFN}(\mathbf{x}) = W_2 \cdot \text{GELU}(W_1 \mathbf{x} + b_1) + b_2,
\end{align}

where the inner dimension is $4H$. GELU~\cite{hendrycks2016gaussian} replaces
ReLU; it avoids the hard zero gate, which tends to improve training for
attention-based models. Both sub-layers use residual connections and dropout.

\subsection{Prediction}

After all $B$ blocks a final layer normalisation is applied. The hidden state
at the last sequence position, $\mathbf{h} \in \mathbb{R}^H$, represents the
user's current preference. Candidate items are scored by dot product:

\begin{align}
  \text{score}(j) = \mathbf{h}^\top \mathbf{e}_j.
\end{align}


\section{Training Setup}

\subsection{Objective}

The model is trained with binary cross-entropy and one sampled negative per
positive. At each non-padding position $t$:

\begin{align}
  \mathcal{L} = -\sum_{t:\, i_{t+1} \neq 0}
    \Bigl[
      \log \sigma(\text{score}(i_{t+1}))
      + \log(1 - \sigma(\text{score}(j_t)))
    \Bigr],
\end{align}

where $j_t$ is drawn uniformly from items the user has not interacted with.

\subsection{Optimisation and Early Stopping}

Adam~\cite{kingma2015adam} is used with learning rate $10^{-3}$ and weight
decay $10^{-5}$. Gradient norms are clipped at 5.0. Training runs for up to
50 epochs; if validation NDCG@10 does not improve for five consecutive epochs,
training stops and the best checkpoint is used for test evaluation.

\subsection{Configurations}

Table~\ref{tab:configs} lists all six configurations. The baseline uses the
smaller value for each dimension. Four ablation settings each change exactly one
dimension while holding the rest at baseline, allowing the contribution of each
dimension to be isolated. The combined configuration sets all four dimensions to
their larger values.

\begin{table}[h]
\centering
\caption{Experimental configurations. Cells that differ from the baseline are
shown in \textbf{bold}.}
\label{tab:configs}
\begin{tabular}{lcccc}
\toprule
Configuration       & $H$  & $B$ & Heads & $L$  \\
\midrule
config\_base         & 64   & 2   & 2     & 50   \\
setting\_num\_blocks  & 64   & \textbf{3}   & 2     & 50   \\
setting\_hidden\_dim  & \textbf{128}  & 2   & 2     & 50   \\
setting\_num\_heads   & 64   & 2   & \textbf{4}     & 50   \\
setting\_max\_seq\_len & 64   & 2   & 2     & \textbf{100}  \\
config\_deeper       & \textbf{128}  & \textbf{3}   & \textbf{4}     & \textbf{100}  \\
\bottomrule
\end{tabular}
\end{table}


\section{Evaluation and Results}

\subsection{Protocol}

Evaluation uses full ranking: the model scores all items not previously seen by
each user, and the held-out target is ranked within that set. All items in a
user's history are masked with $-\infty$ before ranking, except the target
itself, so the model is not penalised for assigning low scores to items it has
never been asked to predict.

Recall@$K$ is 1 if the target appears in the top $K$ positions, else 0.
NDCG@$K$ discounts by rank:

\begin{align}
  \text{NDCG@}K =
  \begin{cases}
    \dfrac{1}{\log_2(\text{rank} + 2)} & \text{if rank} < K, \\
    0 & \text{otherwise,}
  \end{cases}
\end{align}

where rank is 0-based.

\subsection{Results}

Table~\ref{tab:results} reports validation NDCG@10 used for model selection and
test metrics for all six configurations. Five of the six models trained for the
full 50 epochs; \texttt{setting\_num\_heads} was the only one to trigger early
stopping, exiting at epoch 49.

\begin{table*}[t]
\centering
\caption{Full-ranking evaluation on the MovieLens~1M test set.
The best value in each column is \textbf{bold}.
Arrows indicate direction relative to \texttt{config\_base}.}
\label{tab:results}
\begin{tabular}{lccccc}
\toprule
Configuration        & Val NDCG@10 & Recall@10 & Recall@20 & NDCG@10 & NDCG@20 \\
\midrule
config\_base          & 0.0880 & 0.1642 & 0.2504 & 0.0845 & 0.1062 \\
setting\_num\_blocks   & 0.0899 {\small (+2.2\%)} & 0.1672 & 0.2580 & 0.0861 & 0.1090 \\
setting\_hidden\_dim   & 0.1093 {\small (+24.2\%)} & 0.1833 & 0.2743 & 0.0973 & 0.1202 \\
setting\_num\_heads    & 0.0844 {\small ($-$4.1\%)} & 0.1501 & 0.2368 & 0.0760 & 0.0978 \\
setting\_max\_seq\_len  & 0.0957 {\small (+8.8\%)} & 0.1755 & 0.2744 & 0.0905 & 0.1154 \\
config\_deeper        & \textbf{0.1131} {\small (+28.5\%)} & \textbf{0.2007} & \textbf{0.3061} & \textbf{0.1032} & \textbf{0.1298} \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Per-Dimension Analysis}

\textbf{Hidden size} is the most impactful single change. Doubling $H$ from 64
to 128 improves validation NDCG@10 by 24.2\% and test NDCG@10 by 15.2\%. Wider
embeddings give the model more capacity to differentiate items and to encode
complex sequential patterns inside the FFN layers, which scale as $4H$.

\textbf{Sequence length} is the second largest contributor at +8.8\% on
validation NDCG@10. Extending $L$ from 50 to 100 allows the model to condition
on more of each user's history. On MovieLens~1M many users have more than 50
positive ratings, so the additional context is not wasted.

\textbf{Number of blocks} provides a modest but consistent gain of +2.2\%.
Adding a third attention block allows one more layer of non-linear transformation,
but the improvement is small relative to hidden size, suggesting that two blocks
are already sufficient to capture most inter-item dependencies at this scale.

\textbf{Number of heads} is the only dimension where increasing the value hurts
performance, dropping validation NDCG@10 by 4.1\%. The cause is the per-head
dimensionality: at $H = 64$ with four heads, each head operates on a 16-dimensional
subspace. This is likely too small to learn meaningful attention patterns, and
the head projection matrices add parameters without adding useful capacity. The
degradation is clear enough that \texttt{setting\_num\_heads} also triggered
early stopping at epoch 49, unlike every other configuration. When heads are
increased in the combined \texttt{config\_deeper} setting, $H$ is simultaneously
doubled to 128, giving each head a 32-dimensional subspace, which is sufficient
to recover the benefit.

\textbf{Combined configuration.} \texttt{config\_deeper} outperforms the
baseline by 28.5\% on validation NDCG@10, which is larger than any single-
dimension gain. The head dimension interaction above explains part of this: the
gain from adding heads only materialises when $H$ is large enough. The remaining
uplift likely reflects non-additive interactions between depth, width, and
context length that the single-dimension ablations cannot capture.


\section{Discussion}

The controlled ablation yields two results that would not be visible from a
two-configuration comparison. First, hidden size dominates the other three
dimensions at this dataset scale. Second, increasing heads without also
increasing the hidden size makes things worse, not better. Both findings have
direct practical implications: on a compute budget, widening the model is a
safer bet than deepening it or adding heads.

The implementation uses pre-layer normalisation, causal masking, $4\times$ FFN
expansion, and GELU activations. Full-ranking evaluation is used throughout;
sampled evaluation has been shown to produce different rankings than full
evaluation and should not be used for final reporting~\cite{krichene2020sampled}.

Three improvements stand out as the most direct next steps. Training longer is
the clearest one: five of six configurations were still improving at epoch 50,
and a cosine learning rate schedule with a longer budget would likely push
performance up across all settings. Negative sampling is a second avenue;
drawing negatives uniformly means the model rarely sees items close to the
decision boundary, and popularity-weighted or in-batch negatives would sharpen
the gradients. Finally, the head-size interaction suggests that a joint sweep
over $H$ and the number of heads (keeping $H / \text{heads}$ above 32) would
be more informative than sweeping them independently.


\section{Conclusion}

Six SASRec configurations are compared on MovieLens~1M under full ranking. The
ablation shows that hidden size is the most valuable single dimension to scale,
while increasing the number of heads without increasing the hidden dimension
actively hurts performance. The combined configuration achieves Recall@10 of
0.2007 and NDCG@10 of 0.1032. Neither the ablation settings nor the combined
model converged within 50 epochs, so the training budget --- not model capacity
--- appears to be the binding constraint at this scale.


\bibliographystyle{ACM-Reference-Format}
\begin{thebibliography}{9}

\bibitem{kang2018self}
  W.-C. Kang and J. McAuley,
  \textit{Self-Attentive Sequential Recommendation},
  in Proc.\ IEEE ICDM, 2018, pp.\ 197--206.

\bibitem{vaswani2017attention}
  A. Vaswani et al.,
  \textit{Attention Is All You Need},
  in Advances in Neural Information Processing Systems, 2017.

\bibitem{harper2015movielens}
  F.~M. Harper and J.~A. Konstan,
  \textit{The MovieLens Datasets: History and Context},
  ACM Trans.\ Interact.\ Intell.\ Syst., vol.~5, no.~4, pp.~1--19, 2015.

\bibitem{xiong2020layer}
  R. Xiong et al.,
  \textit{On Layer Normalization in the Transformer Architecture},
  in Proc.\ ICML, 2020.

\bibitem{hendrycks2016gaussian}
  D. Hendrycks and K. Gimpel,
  \textit{Gaussian Error Linear Units (GELUs)},
  arXiv:1606.08415, 2016.

\bibitem{kingma2015adam}
  D.~P. Kingma and J. Ba,
  \textit{Adam: A Method for Stochastic Optimization},
  in Proc.\ ICLR, 2015.

\bibitem{krichene2020sampled}
  W. Krichene and S. Rendle,
  \textit{On Sampled Metrics for Item Recommendation},
  in Proc.\ ACM KDD, 2020.

\end{thebibliography}

\end{document}
